Joint Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2024-08-02 src/tex/ms.tex
[showyourwork] overleaf sync

        We present a new method for joint likelihood deconvolution (\jolideconospace) of a set of astronomical observations of the same sky region in the presence of Poisson noise. 
        The observations may be obtained from different instruments with  different resolution, and different point spread functions. \jolideco reconstructs a single flux image by optimizing the posterior distribution based on the joint Poisson likelihood of all observations under a patch-based image prior. The patch prior is parameterised via a Gaussian Mixture model which we train on high-signal-to-noise astronomical images, including data from the  James Webb Telescope and the GLEAM radio survey. This prior favors correlation structures among the reconstructed pixel intensities that are characteristic of those observed in the training images. It is, however, not informative for the mean or scale of the reconstruction. By applying the method to simulated data we show that the combination of multiple observations and the patch-based prior leads to much improved reconstruction quality in many different source scenarios and signal to noise regimes. We demonstrate that with the patch prior \jolideco yields superior reconstruction quality relative to alternative standard methods such as the Richardson-Lucy method. We illustrate the results of \jolideco\ applied to example data from the \chandra X-ray Observatory and the \ferminospace\ Gamma-ray Space Telescope. By comparing the measured width of a counts based and the corresponding \jolideco flux profile of an X-ray filament in SNR {\it 1E 0102.2-7219} we find the deconvolved width of $0.58\pm 0.02$ arcsec to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

@@

    To address the issue with RL decomposing extended emission structures into groups of point sources, multiple additions to the RL method  have been proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparse representations based on \enquote{wavelets} or \enquote{curvelets} \citep{Starck2003}. The challenge is rooted in two fundamentally contradicting goals: a deconvolution process that favors point sources will decompose extend structures present in the image, while introducing local correlations between neighbouring pixels will lead to less sharp reconstruction of point sources.

@@

    \cite{Selig2015A} and \cite{Pumpe2018} approach the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they are able to deconvolve \fermi observations
    into a mixture of point-like and diffuse emission components. While their method works well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it showed artifacts around the Galactic Plane, when applied to gamma-ray data from the Fermi-LAT \cite{Selig2015B}.

@@

    Given a single observation $j$ of an unknown flux image $\mathbf{x}$ and assuming the noise in each pixel $i$ in the recorded counts image $\mathbf{D}_j$ follow independent Poisson distributions with expectation $\boldsymbol{\lambda}_j$, the likelihood $P$ of obtaining the measured image with $N$ pixels is given by:
    %
    \begin{equation}
        \label{eq:poisson}
        P\left( \mathbf{D}_j | \boldsymbol{\lambda}_j \right) = \prod_{i=1}^N \frac{{e^{ - \lambda_{j,i} } \lambda_{i,j} ^ {D_{j,i}}}}{{D_{j,i}!}},
    \end{equation}
    %
    where $\boldsymbol{\lambda}_j$ depends on $\mathbf{x}$ as defined by Equation~\ref{eq:model}. By taking the logarithm, multiplying by $-1$, and dropping the constant terms one can transform the product into a sum over pixels, which is also often called the \enquote{Cash} \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j \right) = \sum_{i=1}^{N_j} \left(\lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}\right),
    \end{equation}
    %  
    By summing over all observations we obtain the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right)
    \end{equation}
    %
    In principle, we could obtain the Maximum Likelihood Estimate (MLE), $\hat{\mathbf{x}}$ of ${\mathbf{x}}$, by directly optimizing Equation~\ref{eq:joint}. Indeed, this is the objective of the RL algorithm.  However, in general this represents an \enquote{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ may look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation
    and the noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by maxima. Depending on the starting value used for $\mathbf{x}$ and the optimization algorithm the computed MLE is likely to be a local maxima. This is fundamentally the reason for RL decomposing estimates into point sources, described in Section~\ref{sec:related}.

    \subsection{A Posteriori Distribution}
    When reconstructing images in the context of inverse problems we are operating in a high dimensional parameter space, because each pixel in the image represents a parameter in the optimization process. By using prior information we can introduce correlations among the parameters, to reduce the \enquote{effective dimensionality} of the problem and guide the optimization process toward unique and more stable solutions.

    Using Bayes' rule we can express the \aposteriori likelihood under a given prior distribution, $P(\mathbf{x} )$:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}.
    \end{equation}
    %
    The Maximum A Posteriori (MAP) estimate of $\mathbf{x}$ is the value that maximizes Equation~\ref{eq:bayes}.
    Taking the logarithm of Equation~\ref{eq:bayes}, replacing the definition of the likelihood $P(\mathbf{D}|\mathbf{x})$ with Equation~\ref{eq:joint} and dropping the normalization term $P(\mathbf{D})$ which is independent of $\mathbf{x}$, we obtain the following expression for the log-posterior distribution
    $\mathcal{L}$:
    %
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left(\mathbf{x} | \mathbf{D} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right) - \beta \cdot \mathcal{P}(\mathbf{x}),
    \end{equation}
    %
    where $\mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right)$ represents the summed log-likelihood for an individual observation $j$ and $\mathcal{P}(\mathbf{x})$ represents the log of the prior distribution with the factor $\beta$ adjusting the weight of the prior relative to the joint log likelihood term. The parameter $\beta$ can be viewed as a hyper-parameter, chosen by the user to adjust the strength of the prior distribution relative to the likelihood. The choice of $\beta$ is optional and for all experiments presented in this paper we set $\beta=1$.

@@

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighbourhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x},
    \end{equation}
    %
    where $\mathbf{P}_n$ represents a neighbourhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixel intensities within the patches using a Gaussian Mixture Model (GMM) \citep{McLachlan2004}. The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete nuisance-parameter that is optimized along with the image $\mathbf{x}$.

    A convenient property of the GMM is that its probability density can be evaluated in closed form. For a given single patch $\mathbf{x}_n$,
    %
    \begin{equation}
        \label{eq:gmm}
        P_{\rm GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2),
    \end{equation}
    %

    Where $\mathcal{N}$ is a multivariate Gaussian normal distribution. The distribution is parameterized by the vector $\mathbf{\mu_k}$, which a dimension equivalent to the number of pixels in a patch and represents the mean of each Gaussian component. The pairwise correlation between pixels is represented by the corresponding covariance matrix $\mathbf{\Sigma}_k$, while the weight of each component is given by $\pi_k$.

    If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimized) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        %\mathcal{P}(\mathbf{x} | \hat{\mathbf{k}})  + \mathcal{P}(\hat{\mathbf{k}}) =
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed and $\pi_{\hat{k}_n}$ the weight of the component $\hat{k}_n$, which maximizes the log-likelihood for the n-th patch.

@@

 \begin{figure}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$~pixels on top of an example image of the size $128 \times 68$~pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the horizontal and vertical \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners. The bottom row shows the red example patch $\mathbf{x}_n$ on the left. The three other patches show the example patch represented by a linear combination of the six largest Eigenvectors of the covariance matrices of the 3 most likely GMM components $k = 105, 77, 17$. This example uses the GMM learned from the JWST Cas A image.
            }
            \label{fig:patches}
            \script{patches.py}
        \end{centering}
    \end{figure}

@@

    Apart from this, the process for choosing the most likely GMM prior component remains unmodified. However, we note that we neglected the prior dependence between neighbouring patches\footnote{That is, we continue to evaluate the log-prior distribution of the full image by summing over the overlapping patches as in Equation~\ref{eq:prior-total}, even though overlapping patches include common pixels and are thus clearly not \aprioir independent.}. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

    \subsubsection{Image Dynamic Range and Patch Normalization}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images. To normalize the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing significantly more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we instead propose a modified normalization for the patches:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}
    %
    By dividing by the mean of the patch, $\Bar{\mathbf{x}}_n$, the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel contrast and not by their overall brightness. This corresponds to the assumption that the structure in the patch is \aprioir independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior distribution of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior distribution is independent of the total flux of the image. The prior distribution simply regularizes the structure in the image, while the overall intensities are typically well constrained by data as quantified by the Poisson likelihood term. Thus we simply assume a uniform prior distribution for the patch intensities. Algorithmically, we replace $\mathbf{x}_n$ with $\tilde{\mathbf{x}}_n$ when evaluating Equations~\ref{eq:gmm} and \ref{eq:khat} and transform $\mathbf{P}_n\mathbf{x}$ via Equation~\ref{eq:patch-norm} when evaluating Equation~\ref{eq:prior-total}.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction was slightly faster with the same reconstruction quality. However we found (a) practical issues with reproducibility of results when using the method on a GPU and (b) a lot of code complexity to handle the edge of images correctly. Give the performance improvement was only small,
    we decided to just stick with the much simpler regular grid and cycle spinning, described in the previous paragraph.

@@

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalization factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \boldsymbol{\lambda}_j = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) + \alpha_j \cdot \mathbf{B}_j
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalization can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

@@

    We also experimented with computing the diagonal of the Hessian of the posterior distribution evaluated at the MAP estimate. However we found this numerically challenging because of the random cycle spinning.

>>>>

        We present a new framework for joint likelihood deconvolution (\jolideconospace) of a set of astronomical observations of the same sky region in the presence of Poisson noise. 
        The observations may be obtained from different instruments with  different resolution, and different point spread functions. \jolideco reconstructs a single flux image by optimizing the posterior distribution based on the joint Poisson likelihood of all observations under one of several prior distributions. Most notably we employ a patch-based image prior, that is parameterised via a Gaussian Mixture model which we train on high-signal-to-noise astronomical images, including data from the James Webb Telescope and the GLEAM radio survey. This prior favors correlation structures among the reconstructed pixel intensities that are characteristic of those observed in the training images. It is, however, not informative for the mean or scale of the reconstruction. By applying the method to simulated data we show that the combination of multiple observations and the patch-based prior leads to much improved reconstruction quality in many different source scenarios and signal to noise regimes. We demonstrate that with the patch prior \jolideco yields superior reconstruction quality relative to alternative standard methods such as the Richardson-Lucy method. We illustrate the results of \jolideco\ applied to example data from the \chandra X-ray Observatory and the \ferminospace\ Gamma-ray Space Telescope. By comparing the measured width of a counts based and the corresponding \jolideco flux profile of an X-ray filament in SNR {\it 1E 0102.2-7219} we find the deconvolved width of $0.58\pm 0.02$ arcsec to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

@@

    To address the issue with RL decomposing extended emission structures into groups of point sources, multiple additions to the RL method  have been proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparse representations based on \enquote{wavelets} or \enquote{curvelets} \citep{Starck2003}. The challenge is rooted in two fundamentally contradicting goals: a deconvolution process that favors point sources will decompose extend structures present in the image, while introducing local correlations between neighboring pixels will lead to less sharp reconstruction of point sources.

@@

    \cite{Selig2015A} and \cite{Pumpe2018} approach the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory (IFT). Based on physical prior assumptions on the spatial distribution of point sources and local correlation structures of diffuse emission, they are able to deconvolve \fermi observations into a mixture of point-like and diffuse emission components. While their method works well for large-scale diffuse emission, such as the Galactic diffuse gamma-ray emission, it shows artifacts around the Galactic Plane, when applied to gamma-ray data from the Fermi-LAT \citep{Selig2015B}.

    \cite{Westerkamp2024} applied the same IFT framework to a Chandra x-ray observation of the supernova remnant (SNR) SN1006. The authors demonstrated that it is possible to separate the point like emission of the extra-galactic background sources from the diffuse foreground emission of the SNR. The method delivered good reconstruction quality on synthetic data. When applied to observational data the authors found plausible reconstructions that include small artifacts, which they attributed to a mismatch between the modelled and actual PSF.

@@

    Given a single observation $j$ of an unknown flux image $\mathbf{x}$ and assuming the noise in each pixel $i$ in the recorded counts image $\mathbf{D}_j$ follow independent Poisson distributions with expectation $\boldsymbol{\lambda}_j$, the likelihood $P$ of obtaining the measured image with $N$ pixels is given by:
    %
    \begin{equation}
        \label{eq:poisson}
        P\left( \mathbf{D}_j | \boldsymbol{\lambda}_j \right) = \prod_{i=1}^N \frac{{e^{ - \lambda_{j,i} } \lambda_{i,j} ^ {D_{j,i}}}}{{D_{j,i}!}},
    \end{equation}
    %
    where $\boldsymbol{\lambda}_j$ depends on $\mathbf{x}$ as defined by Equation~\ref{eq:model}. By taking the logarithm, multiplying by $-1$, and dropping the constant terms one can transform the product into a sum over pixels, which is also often called the \enquote{Cash} \citep{Cash1979} fit statistics:
    %
    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j(\mathbf{x}) \right) = \sum_{i=1}^{N_j} \left(\lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}\right),
    \end{equation}
    %  
    By summing over all observations we obtain the (negative) joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j(\mathbf{x}) \right)
    \end{equation}
    %
    In principle, we could obtain the Maximum Likelihood Estimate (MLE), $\hat{\mathbf{x}}$ of ${\mathbf{x}}$, by directly optimizing Equation~\ref{eq:joint}. Indeed, this is the objective of the RL algorithm.  However, in general this represents an \enquote{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ may look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation
    and the noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by maxima. Depending on the starting value used for $\mathbf{x}$ and the optimization algorithm the computed MLE is likely to be a local maxima. This is fundamentally the reason for RL decomposing estimates into point sources, described in Section~\ref{sec:related}.

    \subsection{A Posteriori Distribution}
    When reconstructing images in the context of inverse problems we are operating in a high dimensional parameter space, because each pixel in the image represents a parameter in the optimization process. High resolution \xray-images are typically composed of  
    %one can expect 
    up to $\mathcal{O}(10^6)$ pixels. By using prior information we can introduce correlations among the parameters, to reduce the \enquote{effective dimensionality} of the problem and guide the optimization process toward unique and more stable solutions.

    Using Bayes' rule we can express the \aposteriori likelihood under a given prior distribution, $P(\mathbf{x} )$:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}.
    \end{equation}
    %
    The Maximum A Posteriori (MAP) estimate of $\mathbf{x}$ is the value that maximizes Equation~\ref{eq:bayes}.
    Taking the logarithm of Equation~\ref{eq:bayes}, replacing the definition of the likelihood $P(\mathbf{D}|\mathbf{x})$ with Equation~\ref{eq:joint} and dropping the normalization term $P(\mathbf{D})$ which is independent of $\mathbf{x}$, we obtain the following expression for the (negative) log-posterior distribution
    $\mathcal{L}$:
    %
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left(\mathbf{x} | \mathbf{D} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j(\mathbf{x}) \right) - \beta \cdot \mathcal{P}(\mathbf{x}),
    \end{equation}
    %
    where $\mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j(\mathbf{x}) \right)$ represents the summed log-likelihood for an individual observation $j$ and $\mathcal{P}(\mathbf{x})$ represents the log of the prior distribution with the factor $\beta$ adjusting the weight of the prior relative to the joint log likelihood term. The parameter $\beta$ can be viewed as a hyper-parameter, chosen by the user to adjust the strength of the prior distribution relative to the likelihood. The choice of $\beta$ is optional and for all experiments presented in this paper we set $\beta=1$.

@@

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighborhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x},
    \end{equation}
    %
    where $\mathbf{P}_n$ represents a neighborhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixel intensities within the patches using a Gaussian Mixture Model (GMM) \citep{McLachlan2004}. The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete nuisance-parameter that is optimized along with the image $\mathbf{x}$.

    A convenient property of the GMM is that its probability density can be evaluated in closed form. For a given single patch $\mathbf{x}_n$,
    %
    \begin{equation}
        \label{eq:gmm}
        P_{\rm GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2),
    \end{equation}
    where $\mathcal{N}$ represents a multivariate Gaussian normal distribution, $\mathbf{\mu_k}$ is the vector of Gaussian means (with dimension equal to the number of pixels in each patch), $\mathbf{\Sigma}_k$ is the variance-covariance matrix of the Gaussian distribution (which tabulates the variances and pairwise covariances of the patch pixels), and  $\pi_k$ is the weight of component $k$ of the GMM.

    If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimized) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ and $\pi_{\hat{k}_n}$ is the weight of GMM component $\hat{k}_n$, with $\hat{k}_n$ being the index of the GMM component that maximizes the log-prior distribution for patch $n$.

@@

 \begin{figure}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$~pixels on top of an example image of the size $128 \times 68$~pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the horizontal and vertical \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners. The bottom row shows the red example patch $\mathbf{x}_n$ on the left. The three other patches show this example patch represented by a linear combination of the six largest Eigenvectors of the covariance matrices of the 3 most likely GMM components $k = 105, 77, 17$. This example uses the GMM learned from the JWST Cas A image.
            }
            \label{fig:patches}
            \script{patches.py}
        \end{centering}
    \end{figure}

@@

    Apart from this, the process for choosing the most likely GMM prior component remains unmodified. However, we note that we neglected the prior dependence between neighboring patches\footnote{That is, we continue to evaluate the log-prior distribution of the full image by summing over the overlapping patches as in Equation~\ref{eq:prior-total}, even though overlapping patches include common pixels and are thus clearly not \aprioir independent.}. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

    \subsubsection{Image Dynamic Range and Patch Normalization}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images. To normalize the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing significantly more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we instead propose a modified normalization for the patches:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \widetilde{\mathbf{P}_n\mathbf{x}}
        = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}.
    \end{equation}
    %
    By dividing by the mean of the patch, $\Bar{\mathbf{x}}_n$, the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel contrast and not by their overall brightness. This corresponds to the assumption that the structure in the patch is \aprioir independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior distribution of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior distribution is independent of the total flux of the image. The prior distribution simply regularizes the structure in the image, while the overall intensities are typically well constrained by data as quantified by the Poisson likelihood term.

@@

    A variation of cycle spinning specifically for the patch prior is proposed in \cite{Parameswaran2018}. Instead of a regular grid for the patches they propose to work with a randomised grid of patches, where the center of each patch is randomly shifted, while ensuring that patches overlap by at least one pixel. In doing so they find that a smaller total number of patches can be used, while retaining the same reconstruction quality. This improves the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction was slightly faster with the same reconstruction quality. We found, however,  (a) there are practical issues with reproducibility of results when using this method on a GPU and (b) the method requires significant code complexity to handle the edge of images correctly. Given that the performance improvement was small,
    we thus decided to use the simpler regular grid and cycle spinning, as described in the previous paragraph.

@@

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalization factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \begin{aligned}
        \boldsymbol{\lambda}_j(\mathbf{x}, \delta_{x,j}, \delta_{y,j}, \alpha_j) = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) \\ + \alpha_j \cdot \mathbf{B}_j
        \end{aligned}
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalization can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

    Modifying the (negative) log-posterior distribution given in Equation~\ref{eq:total} for the normalization of the $\mathbf{x}_n$ in Equation~\ref{eq:patch-norm}  and the adjustment for systematic effects in Equation~\ref{eq:model-npred-calibration} yields,
    \begin{equation}
        \label{eq:total-final}
        \begin{aligned}
        \mathcal{L}\left(\mathbf{x}, \mathbf{k}, \boldsymbol{\delta}, \boldsymbol{\alpha} | \mathbf{D} \right) 
        = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \boldsymbol{\lambda}_j(\mathbf{x}, \delta_{x,j}, \delta_{y,j},\alpha_j) \right) 
        \\ - \beta \left\{
         \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\widetilde{\mathbf{P}_n \mathbf{x}} | {k}_n) +
        \sum_{n = 1}^N \log \pi_{{k}_n} \right\},
        \end{aligned}
    \end{equation}
    where $\mathbf{k} = (k_1, \ldots k_N)$, $\boldsymbol{\delta} = (\delta_{x,1},\delta_{y,1},\ldots, \delta_{x,J},\delta_{y,J})$ and $\boldsymbol{\alpha} = (\alpha_1,\ldots, \alpha_J)$. For fixed $\mathbf{x}$, Equation~\ref{eq:total-final} is optimized as a function of $\mathbf{k}$ by Equation~\ref{eq:khat}, but with $\mathbf{x}_n$ replaced by $\widetilde{\mathbf{P}_n \mathbf{x}}$.  The joint prior distribution of $(\mathbf{x}, \mathbf{k}, \boldsymbol{\delta}, \boldsymbol{\alpha})$ is given by the term in Equation~\ref{eq:total-final} within curly brackets. Note that this places uniform prior distributions on $\boldsymbol{\delta}$ and $\boldsymbol{\alpha}$ and formally sets the prior on $\mathbf{P}_n\mathbf{x}$ to the GMM evaluated at  $\widetilde{\mathbf{P}_n \mathbf{x}}$. Since this does not depend on $\mathbf{\bar x}_n$, we implicitly assume a uniform prior on each $\mathbf{\bar x}_n$. In practice, we do not expect the choice of prior distribution on the scale of each patch to influence the final fit, but some users may prefer to assume a flat prior on the log of each $\mathbf{\bar x}_n$ by adding $\sum_{n=1}^N \mathbf{\bar{x}}_n$ within the curly brackets on the right-hand-side of Equation~\ref{eq:total-final}.

@@

    How well these bootstrap error estimates approximate the true posterior uncertainty and/or the frequency coverage of the derived bootstrap intervals is a topic of future study. 
    We also experimented with computing the diagonal of the Hessian of the posterior distribution evaluated at the MAP estimate. However we found this numerically challenging because of the random cycle spinning.

    \section{Implementation}
    \subsection{\jolideco Framework}
    The goal of the reconstruction process is to find the mode of the posterior distribution defined by Equation~\ref{eq:total-final}. Given that there is a parameter for each pixel in the image, this is a high dimensional optimization problem. Differentiable programming frameworks such as PyTorch~\citep{Pytorch2019} allow for solving high dimensional optimization problems using back-propagation. They also offer a variety of optimization methods such as stochastic gradient decent (SGD). The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of the observations represents a \enquote{mini batch} in terms of ML.
